---
title: "CAS Data Summer Camp"
output: html_notebook
---
## Session One Day Five
Today we are going to do more work with data files as well as modeling and visualizing data.

### Data Handling
```{r}
rm(list=ls())
df <- read.csv("salaries.csv", header = T)
head(df)

colnames(df) <- c("Rk","Player","Tm","x25","x26","x27","x28","x29","x30","Guaranteed")

head(df)

typeof(df$x25)

df$x25 <- gsub(",","",df$x25)

head(df$x25)

df$x25 <- gsub("$","",df$x25)

head(df$x25)

df$x25 <- gsub("\\$","",df$x25)

df$x25

head(df$x26)

df$x26 <- gsub("$","",df$x26, fixed = T)

head(df$x25)

typeof(df$x25)

df$x25 <- as.numeric(df$x25)

head(df$x25)

typeof(df$x25)
```

Batch processing
```{r}
rm(list=ls())
df <- read.csv("salaries.csv", header = T)
colnames(df) <- c("Rk","Player","Tm","x25","x26","x27","x28","x29","x30","Guaranteed")
head(df)

colnames(df)

salCols <- c("x25","x26","x27","x28","x29","Guaranteed")

for (i in salCols) {
  df[,i] <- gsub(",","",df[,i])
  df[,i] <- gsub("$","",df[,i], fixed = T)
  df[,i] <- as.numeric(df[,i])
}
head(df)

summary(df)

for (i in salCols) {
  print(paste("Average salary in",i,"is",round(mean(df[,i], na.rm = T), 2)))
}
```

### Combining Data
There are generally speaking 3 ways to combine data in R. cbind, rbind, and merge. cbind puts datasets next to one another. You need to be extremely careful with cbind because it defaults to fill data from both binded datasets.

```{r}
rm(list=ls())
df1 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5)
df2 <- data.frame(Name = c('c1', 'c2', 'c3', 'd1', 'd2', 'd3'), data2 = 6:11)

df3 <- cbind(df1,df2)

df4 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5)
df5 <- data.frame(Name = c('c1', 'c2', 'c3'), data2 = 6:8)

df6 <- cbind(df4,df5)
```

rbind just puts datasets on top of one another. Columns must match
```{r}
rm(list=ls())
df1 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5)
df2 <- data.frame(Name = c('c1', 'c2', 'c3', 'd1', 'd2', 'd3'), data1 = 6:11)

df3 <- rbind(df1,df2)

df4 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5)
df5 <- data.frame(Name = c('c1', 'c2', 'c3'), data1 = 6:8)

df6 <- rbind(df4,df5)

df7 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5,data2 = 54:59)
df8 <- data.frame(Name = c('c1', 'c2', 'c3', 'd1', 'd2', 'd3'), data1 = 6:11)

df10 <- rbind(df7,df8)
```

Merging requires a shared columnn name or you can define a column you want to merge on.
```{r}
rm(list=ls())
df1 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5)
df2 <- data.frame(Name = c('a1', 'a2', 'a3', 'c1', 'c2', 'c3'), data2 = 6:11)

df3 <- merge(df1,df2)

df4 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5)
df5 <- data.frame(Name = c('a1', 'a2', 'a3'), data2 = 6:8)

df6 <- merge(df4,df5)

df7 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5, data2 = 6:11, data3 = 12:17)
df8 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5, data2 = 18:23, data3 = 12:17)

df9 <- merge(df7,df8)

df9 <- merge(df7,df8, by = "Name")

df7 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5, data2 = 6:11, data3 = 12:17)
df8 <- data.frame(ID = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data4 = 0:5, data5 = 18:23, data6 = 12:17)

df9 <- merge(df7,df8, by.x = "Name", by.y = "ID")

df7 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5, data2 = 6:11, data3 = 12:17)
df8 <- data.frame(ID = c('a1', 'a2', 'a3', 'c1', 'c2', 'b3'), data4 = 0:5, data5 = 18:23, data6 = 12:17)

df9 <- merge(df7,df8, by.x = "Name", by.y = "ID")

df7 <- data.frame(Name = c('a1', 'a2', 'a3', 'b1', 'b2', 'b3'), data1 = 0:5, data2 = 6:11, data3 = 12:17)
df8 <- data.frame(ID = c('a1', 'a2', 'a3', 'c1', 'c2', 'b3'), data4 = 0:5, data5 = 18:23, data6 = 12:17)

df9 <- merge(df7,df8, by.x = "Name", by.y = "ID", all = T)
```

### Grouping
We can use the aggregate feature to group our data into smaller chunks.
```{r}
rm(list=ls())
df <- read.csv("salaries.csv", header = T)
colnames(df) <- c("Rk","Player","Tm","x25","x26","x27","x28","x29","x30","Guaranteed")
salCols <- c("x25","x26","x27","x28","x29","Guaranteed")
for (i in salCols) {
  df[,i] <- gsub(",","",df[,i])
  df[,i] <- gsub("$","",df[,i], fixed = T)
  df[,i] <- as.numeric(df[,i])
}

a <- aggregate(Guaranteed ~ Tm, data = df, FUN = "mean")

mean(df$Guaranteed[df$Tm=="DET"])

b <- aggregate(Guaranteed ~ Tm, data = df, FUN = "sum")

c <- aggregate(x25 ~ Tm, data = df, FUN = "mean")

d <- aggregate(Player ~ Tm, data = df, FUN = "length")

```

### Modeling in R
Modeling in R is also very straightforward for simple models.
```{r}
rm(list=ls())
A <- read.csv("sav.csv", sep = ',', header = T)
A[A==99] <- NA
summary(A)

lm(payatt ~ changemind, data = A)

summary(lm(payatt ~ changemind, data = A))

summary(lm(engagepol ~ seepol, data = A))

model1 <- lm(payatt ~ changemind + seepol, data = A)

summary(model1)

model1$coefficients

model2 <- lm(A$payatt ~ A$changemind * A$seepol)

summary(model2)
```

We can also do things like categorical modeling or logistic regression.
```{r}
letters_set <- c("a","b","c","d")
value_set <- c(1,2,3,4)
random_letters <- sample(letters_set, 50, replace = TRUE)
random_values <- sample(value_set, 50, replace = TRUE)
data <- data.frame(id = random_letters, score = random_values)
data$id <- as.factor(data$id)
head(data)

summary(lm(score ~ id, data = data))

levels(data$id)

data$id <- relevel(data$id, ref = "c")

summary(lm(score ~ id, data = data))
```

```{r}
rm(list=ls())
score_set <- runif(500, min = 1, max = 5)
value_set <- c(0,1)
random_values <- sample(value_set, 500, replace = TRUE)
data <- data.frame(value = random_values, score = score_set)
head(data)

summary(glm(value ~ score, data = data, family = "binomial"))
```

### Showing Our Models
We can use a package, something we haven't had to do yet, to get our models out into workable text.
```{r}
install.packages("modelsummary")
install.packages("pandoc")
library(modelsummary)

rm(list=ls())
A <- read.csv("sav.csv", sep = ',', header = T)
A[A==99] <- NA
summary(A)

# Save your models as objects
m1 <- lm(payatt ~ changemind, data = A)
m2 <- lm(payatt ~ seepol, data = A)
m3 <- lm(payatt ~ engagepol, data = A)

# Create a list of your models with labels for each one
models <- list(
  "Change Minds" = m1,
  "See Politics" = m2,
  "Engage Politics" = m3
)

models
```

```{r}
modelsummary(models)

modelsummary(models, estimate = "{estimate} {stars}")

modelsummary(models, estimate = "{estimate} [{conf.low}, {conf.high}, {stars}]")

modelsummary(models, estimate = "{estimate} {stars}", coef_rename = c("changemind" = "Change Mind", "seepol" = "See Politics"))

modelsummary(models, estimate = "{estimate} {stars}", coef_rename = c("changemind" = "Change Mind", "seepol" = "See Politics", "engagepol" = "Engage Politics"), output = "models.docx")
```


### Plotting
There is a lot you can do with plotting. This is a basic overview of the options.
```{r}
rm(list=ls())
df <- read.csv("caremissions.csv", sep = ',', header = T)

plot(df$ob.fuel,df$wl.fuel)

# Markdown or Notebooks in R are a little different from plots because you need to do everything all at once.
plot(df$ob.fuel,df$wl.fuel)
title("Onboard Fuel Economy vs. Tested Fuel Economy")

# We can get around this by nesting our plot calls inside of a {}
{
  plot(df$ob.fuel,df$wl.fuel)
  title("Onboard Fuel Economy vs. Tested Fuel Economy")
}

```

```{r}
{
  plot(df$ob.fuel, df$wl.fuel, xlab = "Onboard Fuel", ylab = "Measured Fuel")
  title("Onboard Fuel Economy vs. Tested Fuel Economy")
}

summary(df$ob.fuel)
summary(df$wl.fuel)

{
  plot(df$ob.fuel, df$wl.fuel, xlab = "Onboard Fuel", ylab = "Measured Fuel", axes = F)
  title("Onboard Fuel Economy vs. Tested Fuel Economy")
  axis(1, at = seq(0,22, by = 2))
  axis(2, at = seq(0,16, by = 2), las = 1)
}

{
  plot(df$ob.fuel, df$wl.fuel, xlab = "Onboard Fuel", ylab = "Measured Fuel", axes = F, xlim = c(0,22), ylim = c(0,16))
  title("Onboard Fuel Economy vs. Tested Fuel Economy")
  axis(1, at = seq(0,22, by = 2))
  axis(2, at = seq(0,16, by = 2), las = 1)
}
```

### Text Modeling

```{r}
library(stringr)
rm(list=ls())

A <- read.csv("transcripts.csv")
summary(A)
table(A$president)
```

We need to be careful how we ask R to look at text.
```{r}
A$apple1 <- grepl("apple", A$transcript)
table(A$apple1)

A$apple2 <- grepl("apple", A$transcript, ignore.case = TRUE)
table(A$apple2)

table(A$apple1,A$apple2)
```

```{r}
A$freedom <- str_count(A$transcript, "freedom")
A$liberty <- str_count(A$transcript, "liberty")
A$rights <- str_count(A$transcript, "rights")
A$military <- str_count(A$transcript, "military")

sum(A$freedom,A$liberty,A$rights,A$military)

countwords <- c('freedom','liberty','rights',"military")

A$count <- 0
A$patriotismWords <- 0

for (i in countwords) { 
  A$count <- str_count(A$transcript, i)
  A$patriotismWords <- A$patriotismWords + A$count
  #print(paste("There are:",A$count,"instances of",i,"and the total words so far is:",A$patriotismWords,"in that row.", sep = " "))
  A$count <- 0
}
table(A$patriotismWords)

sum(A$patriotismWords)

# This is another way to remove objects, just one at a time or with a list.
rm(i)
rm(countwords)
```

### Modeling Patriotism Words
```{r}
summary(lm(patriotismWords ~ president, data = A))

A$nums <- length(A$date):1 # this creates a chronological numbering system with row 1 starting at 244 and working its way down to 1.

summary(lm(patriotismWords ~ nums, data = A))
```

```{r}
A$year <- as.numeric(format(as.Date(A$date, format="%Y-%m-%d"),"%Y"))

table(A$year)

summary(lm(patriotismWords ~ year, data = A))
```

```{r}
B <- read.csv("presidentparty.csv")
head(B)

C <- merge(A, B, by.x = "president",by.y = "President", all.x = T)

table(C$Party, useNA = "always")

table(B$Party, useNA = "always")

B$Party <- trimws(B$Party, which = "both")
?trimws
table(B$Party, useNA = "always")
```

```{r}
table(C$president[is.na(C$Party)])

table(B$President)

B$President[B$President=="Bill Clinton"] <- "William J. Clinton"

B$President[B$President=="George H. W. Bush"] <- "George Bush"
B$President[B$President=="Gerald Ford"] <- "Gerald R. Ford"
B$President[B$President=="Martin Van Buren"] <- "Martin van Buren"

C <- merge(A, B, by.x = "president",by.y = "President", all.x = T)

table(C$Party, useNA = "always")

table(C$president[is.na(C$Party)])

C$Party[C$president=="Donald J. Trump"] <- "Republican"

table(C$Party, useNA = "always")
```

```{r}
summary(lm(patriotismWords ~ Party, data = C))

C$Party <- as.factor(C$Party)

C$Party <- relevel(C$Party, ref = "Democratic-Republican")

summary(lm(patriotismWords ~ Party, data = C))

t.test(C$patriotismWords[C$Party=="Democratic"],C$patriotismWords[C$Party=="Republican"])
```


```{r}
plot(C$nums,C$patriotismWords)

plot(C$year,C$patriotismWords, type = "l")

C <- C[order(C$date),]

plot(C$year,C$patriotismWords, type = "l")
```

```{r}

# We could add some vertical lines where the US declared war that year to see if there might be a relationship.
{
  plot(C$year, C$patriotismWords, type = "l")
abline(v = 1812)
abline(v = 1846)
abline(v = 1898)
abline(v = 1898)
abline(v = 1941)
abline(v = 1990)
}

{
  plot(C$year, C$patriotismWords,
     type = "l",
     col = "dodgerblue",
     lwd = 2
)
}

{
  plot(C$year, C$patriotismWords,
     type = "l",
     col = "dodgerblue",
     lwd = 2,
     las = 1,
     main="Percentage of Patriotism Words in SOTU", 
     xlab="Year of Speech",
     ylab=""
)
}

summary(C$year)
summary(C$patriotismWords)
{
  plot(C$year, C$patriotismWords,
     type = "l",
     col = "dodgerblue",
     lwd = 2,
     las = 1,
     main="Percentage of Patriotism Words in SOTU", 
     xlab="Year of Speech",
     ylab="",
     axes = F
)
axis(1, at=seq(1790,2018,by=8), las = 2)
axis(2, at=seq(0,150,by=25),las=1)
abline(lm(C$patriotismWords ~ C$year), 
       col = "black", 
       lwd = 2
)
}

C$color <- "darkgray"
C$color[C$Party=="Democratic"] <- "dodgerblue"
C$color[C$Party=="Republican"] <- "firebrick"

table(C$Party,C$color)

{
plot(C$year, C$patriotismWords,
     type = "p",
     pch = 18,
     col = C$color,
     lwd = 2,
     las = 1,
     main="Percentage of Patriotism Words in SOTU", 
     xlab="Year of Speech",
     ylab="",
     axes = F
)
  axis(1, at=seq(1790,2018,by=8), las = 2)
  axis(2, at=seq(0,150,by=25),las=1)
  abline(lm(C$patriotismWords ~ C$year), 
       col = "black", 
       lwd = 2
)
}
```

```{r}
# There are a number of different kinds of visuals we can use with our data.
rm(list=ls())
# Load in the Snapchat 2022 political ads database and get a basic understanding of what is in there.
A <- read.csv("PoliticalAds.csv")
head(A)
summary(A)
```
```{r}
# There is likely a relationship between how much a company spends and how many impressions they get, so we will check that with a cor.test()
cor.test(A$Spend,A$Impressions)
```

```{r}
# We can look at those variables as histograms and barplots. Histograms show the distribution of data into consistent bins, or categories. Barplots show the distribution of all of the data. For data that is skewed right or left, histograms can quickly show that because higher frequencies will be to the right(left).
hist(A$Spend)
hist(A$Spend, breaks = 100)
hist(A$Impressions)

barplot(A$Spend)
barplot(A$Impressions)
```

```{r}
# A basic plot we can do is a scatter plot. We will first just plot the spend variable and then the impressions variable
plot(A$Spend)
plot(A$Impressions)

# We can then look at the relationship between these two variables in a plot
plot(A$Spend,A$Impressions)
```

```{r}
{
plot(A$Spend, A$Impressions,
     col = ifelse(A$CountryCode=="united states",'blue','red'),
     las = 1,
     pch = 20,
     main="Relationship between Ad Spending and Impressions", 
     xlab="",
     ylab="",
     axes = F,
     ylim = c(0,25000000),
     xlim = c(0,200000)
)
summary(A$Spend)
summary(A$Impressions)
axis(1, at=seq(0,200000,by=20000), las = 2)
axis(2, at=seq(0,25000000,by=2500000),las=1)
}
```


```{r}
# Looks a bit better but also has some problems with the margins. We've added some color!
{
  plot(A$Spend, A$Impressions,
     col = ifelse(A$CountryCode=="united states",'blue','red'),
     las = 1,
     pch = 20,
     main="Relationship between Ad Spending and Impressions", 
     xlab="",
     ylab="",
     axes = F,
     ylim = c(0,25000000),
     xlim = c(0,200000)
)
labX <- c("$0","$20,000","$40,000","$60,000","$80,000","$100,000","$120,000","$140,000","$160,000","$180,000","$200,000")
labY <- c("0","2,500,000","5,000,000","7,500,000","10,000,000","12,500,000","15,000,000","17,500,000","20,000,000","22,500,000","25,000,000")
axis(1, at=seq(0,200000,by=20000), labels = labX, las = 2)
axis(2, at=seq(0,25000000,by=2500000), labels = labY, las = 1)
}
```


```{r}
# Margins are fixed and we can save the plot as a pdf. Also throw in a regression line.
{
  pdf(file = "snapchatplot.pdf")
par(mar=c(6,6,4,2.1)) #bottom, left, right, top and their defaults are 5.1,4.1,4.1,2.1
par(family = "serif")
plot(A$Spend, A$Impressions,
     col = ifelse(A$CountryCode=="united states",'blue','red'),
     las = 1,
     pch = 20,
     main="Relationship between Ad Spending and Impressions", 
     xlab="",
     ylab="",
     axes = F,
     ylim = c(0,25000000),
     xlim = c(0,200000)
)
labX <- c("$0","$20,000","$40,000","$60,000","$80,000","$100,000","$120,000","$140,000","$160,000","$180,000","$200,000")
labY <- c("0","2,500,000","5,000,000","7,500,000","10,000,000","12,500,000","15,000,000","17,500,000","20,000,000","22,500,000","25,000,000")
axis(1, at=seq(0,200000,by=20000), labels = labX, las = 2)
axis(2, at=seq(0,25000000,by=2500000), labels = labY, las = 1)
abline(lm(A$Impressions ~ A$Spend), 
       col = "black", 
       lwd = 2
)
dev.off()
}
```

# Task 1: Sentiment Analysis and Visualizing Sentiment Scores 
In this notebook, we explore advanced text manipulation and analysis techniques using R. Our focus is on utilizing packages such as stringi, stringr, and quanteda to process large text datasets. We will demonstrate how to clean and prepare text data, perform sentiment analysis, and visualize the results effectively. This session is designed to provide hands-on experience with real-world data, helping you to develop skills in managing and interpreting textual information.

###Setup

```{r}
# Clearing the workspace
rm(list=ls())

# Installing necessary packages
install.packages("stringi")
install.packages("stringr")
install.packages("quanteda")

# Loading the installed packages
library(stringi)
library(stringr)
library(quanteda)
```

### Data Loading and Preparation

```{r}
# Reading the data
A <- read.csv("transcripts.csv")
A$year <- as.numeric(format(as.Date(A$date, format="%Y-%m-%d"),"%Y"))
summary(A)
```


### Text Data Manipulation

Next, we will split a large text into smaller parts and analyze it.

```{r}
# Preparing data frame for manipulation
A$caseid <- as.numeric(rownames(A))
caseid <- A$caseid

# Creating an empty data frame for results
df <- data.frame(matrix(ncol = 4, nrow = 0))
colnames(df) <- c('text', 'year','pos','president')

# Loop to split text into paragraphs and store in dataframe
for (i in caseid) {
  x <- strsplit(A$transcript[A$caseid==i], split = "\n")
  y <- unlist(x)
  y <- as.data.frame(y)
  colnames(y) <- "text"
  y$year <- A$year[A$caseid==i]
  y$president <- A$president[A$caseid==i]
  y$pos <- 1:nrow(y)
  df <- rbind(df,y)
}
```

```{r}
# Preparing for sentiment analysis using an existing dictionary
# Note: The quanteda package provides multiple dictionary options

# Reassign caseid since we have a new df
df$caseid <- as.numeric(rownames(df))

df$wc <- stri_count(df$text, regex = "\\W+")
df <- df[df$wc > 3,]
docs <- df[, c("caseid", "text")]

# Checking paragraphs for potential formatting issues by word count
length(df$year[df$wc < 50])

# Creating columns to store sentiment scores
df$negative <- NA
df$positive <- NA

# Analyzing text data in batches to manage memory and processing time effectively
n <- 1000
df$batch <- ceiling(as.numeric(rownames(df)) * (1/n))
print(paste0("Number of Batches: ", max(df$batch)))

# This runs the actual LSD sentiment dictionary on the corpus. 

for (i in unique(df$batch)) {
  docs <- df[df$batch == i, c("batch", "caseid", "text")]
  print(paste0("Running Batch: ", i))
  lsd <- dfm(tokens_lookup(tokens(docs$text), dictionary = data_dictionary_LSD2015, exclusive = TRUE))
  lsd <- convert(lsd, to = "data.frame")
  df$negative[df$batch == i] <- lsd$negative
  df$positive[df$batch == i] <- lsd$positive
  print(paste0("Completed Batch: ", i))
  rm(lsd)
}

# Calculating the overall sentiment scores and displaying the distribution
df$lsd <- ((df$positive - df$negative) / df$wc) * 100
table(df$lsd)
plot(df$lsd)

```
```{r}
# What about aggregation and plotting?
year <- aggregate(lsd ~ year, data = df, FUN = "mean")
pres <- aggregate(lsd ~ president, data = df, FUN = "mean")

plot(year$year,year$lsd)

{
  plot(year$year,year$lsd, type = "l")
abline(v=1950)
}

plot(pres$president,pres$lsd)
```

```{r}

# But we get an error for this! It doesn't know how to handle the president names. We can solve that by assigning them numeric labels and then changing labels in the plot.
pres$num <- 1:nrow(pres)

plot(pres$num,pres$lsd)

# We will also drop the first names of the presidents. This code keeps JUST the last word in the string.
pres$first <- word(pres$president,-1)
table(pres$first)

#This isn't ideal since there are repeat names. We could do a custom list, though. This is tricky for a lot of observations but we don't have too many.
pres$president[pres$president=="George W. Bush"] <- "George WBush"
pres$president[pres$president=="Franklin D. Roosevelt"] <- "FDR"
pres$president[pres$president=="John Quincy Adams"] <- "John QAdams"
pres$president[pres$president=="Lyndon B. Johnson"] <- "LBJ"
pres$president[pres$president=="Martin van Buren"] <- "MVB"
pres$first <- word(pres$president,-1)
table(pres$first)

# Now plot the whole thing
{
  par(mar=c(6,4.1,4.1,2.1))
plot(pres$num,pres$lsd, type = "b", axes = F, ylab = "", xlab = "", ylim = c(0,6))
axis(1, at=seq(1,42,by=1), labels = pres$first, las = 2)
summary(pres$lsd)
axis(2, at=seq(0,6,by=1), las = 1)
}
```